---
title: "Movie Ratings"
author: "David Melchor"
format: 
  html:
    embed-resources: true
    toc: true
editor: visual
---

## Introduction

The objective of this assignment was to collect simple movie-rating data, store it in a SQL database, and analyze it in R. I created a WhatsApp poll to survey members of my family, asking them to rate the following movies on a 1–5 scale (with 1 being bad and 5 being excellent): *Prince of Egypt*, *Coco*, *Encanto*, *The Lion King*, *Moana*, and *Spider-Man: Into the Spider-Verse*. Participants were instructed to rate only the movies they had seen and to leave any unseen movies unrated. In total, I received 14 responses.

## Loading Packages in R

```{r}
#| label: packages

# Load PostgreSQL packages
pacman::p_load(DBI, RPostgres, tidyverse, knitr, kableExtra)
```

## Connecting to SQL Server

```{r}
#| label: connect

# Connect to database
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "postgres",
  host = "localhost",
  port = 5432,
  user = Sys.getenv("USER")
  )

# Test connection
dbIsValid(con)
```

## Creating a Data Table

I made my job a bit harder by using a WhatsApp poll, as I had to manually populate the SQL database rather than simply exporting responses from Google Forms into a .csv file. However, manually building the table in the script ensures the entire document is fully self-contained and reproducible—anyone can run this code and obtain the exact same results.

```{r}
#| label: create-table

# Drop table to avoid duplicate rows
dbExecute(con, "
  DROP TABLE IF EXISTS movie_ratings;
")

# Create SQL data table "movie ratings"
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS movie_ratings (
    person_id SERIAL PRIMARY KEY,
    person_name VARCHAR(50),
    prince_of_egypt NUMERIC(2,1),
    coco NUMERIC(2,1),
    encanto NUMERIC(2,1),
    lion_king NUMERIC(2,1),
    moana NUMERIC(2,1),
    into_the_spider_verse NUMERIC(2,1)
  );
")
```

```{r}
#| label: insert-data

# Insert survey responses to table
dbExecute(con, "
  INSERT INTO movie_ratings (
    person_name,
    prince_of_egypt,
    coco,
    encanto,
    lion_king,
    moana,
    into_the_spider_verse)
  VALUES
    ('Alain', 5, 5, 2, 5, 3, 4),
    ('Liliana', 5, 3, 2, 5, 3, 3),
    ('Maya', 5, 3, 3, 4, 3, 4),
    ('Vanesa', 5, 4, 2, 4, 5, 5),
    ('Heather', 5, 5, 3, 5, 3, 4),
    ('Elisa', 5, 5, 3, 5, 3, 4),
    ('Nate', 3, 5, NULL, 4, 4, NULL),
    ('Jeff', 4, 4, 3, 5, 3, NULL),
    ('Steve', NULL, 5, NULL, 3, 3, NULL),
    ('Rachel', 5, 3, NULL, 4, 2, NULL),
    ('Stewart', 5, 3, NULL, 4, 2, NULL),
    ('Becca', NULL, 5, 3, 4, 2, NULL),
    ('Melanie', NULL, 5, 4, 3, 5, NULL),
    ('Mateo', 5, 4, 3, 4, NULL, 5);
")
```

## Lodading the Data Into R

```{r}
#| label: load-to-R

# Move the data table movie_ratings to R
ratings <- dbGetQuery(con,
                "SELECT * FROM movie_ratings;")

# Inspect the data
glimpse(ratings)
```

## Manipulating, Reshaping and Cleaning the Data

After inspecting the dataset, I initially struggled to generate summary statistics on the wide-format table. I realized I needed to pivot the data into a long format, creating two new variables: `movie` and `rating`. Reshaping the data allowed me to group by `movie` and compute the required summary metrics.

```{r}
#| label: pivot-longer

# Pivot data longer to group_by() movies
ratings <- ratings |> 
  pivot_longer(
    cols = -c(person_id, person_name),
    names_to = "movie",
    values_to = "rating"
  )
```

```{r}
#| label: re-label-titles

# Re-label titles
ratings <- ratings |> 
  mutate(
    movie = case_when(
      movie == "prince_of_egypt" ~ "Prince of Egypt",
      movie == "coco" ~ "Coco",
      movie == "encanto" ~ "Encanto",
      movie == "lion_king" ~ "The Lion King",
      movie == "moana" ~ "Moana",
      movie == "into_the_spider_verse" ~ "Spider-Man: Into the Spider-Verse",
      TRUE ~ "movie"
    )
  )
```

```{r}
#| label: avg-rating

# Create summary table
movie_summary <- ratings |> 
  group_by(movie) |> 
  summarise(
    times_rated = sum(!is.na(rating)),
    missing_count = sum(is.na(rating)),
    mean_rating = round(mean(rating, na.rm = TRUE), 2),
  ) |> 
  arrange(desc(mean_rating))
```

## Formatting for Presentation

Once I had `movie_summary` created, I was able to turn this data frame into a table using the packages `knitr` and `kableExtra`

```{r}
#| label: view-table

# Make summary table
movie_summary |> 
  kbl(
    caption = "Table 1: Overall Movie Ratings Summary",
    col.names = c("Movie Title", "Times Rated", "Missing Count", "Mean Rating"),
    align = c("l", "c", "c", "c")
  ) |> 
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) |> 
  column_spec(1, bold = TRUE) %>%
  column_spec(4, bold = TRUE, color = "#2A6496")
```

## Conclusion

Although I assumed I had chosen universally popular movies, I was surprised to find several missing values where participants had not seen certain films. A primary challenge in this assignment was creating and populating the SQL table from scratch; while I have basic familiarity with SQL, I had not previously constructed a table and inserted data directly. Performing this step directly in R would have been simpler, but using SQL fulfilled a key assignment requirement. In order to get help with this, I referred to the "Practical SQL A Beginner's Guide to Story Telling with Data" ebook found on this [site](https://www.kea.nu/files/textbooks/humblelearn2code/practicalsql.pdf). Chapter 1 had instructions on how to create a table and insert data manually.

I also noticed that every time I ran the code, the number of times the movies were rated kept increasing in my data table. I didn't realize that every time I rendered or ran the code, I was adding duplicate rows. I learned that I needed to add the `DROP TABLE IF EXISTS movie_ratings;` command right before creating the table so it would reset cleanly and stop that behavior.

Other challenges I faced were understanding the need to pivot the data into a long format so I could group it by movie and run summary statistics to do the analysis. That was the real key to getting the results I needed. It was also a good learning experience to use `knitr` and `kableExtra` to format my table, since using `View(movie_summary)` didn't let the document render as I expected. I found this [video](https://www.youtube.com/watch?v=JqUViTDoSEo) on YouTube that explains how to use the `knitr` and `kableExtra` packages to create the table I did.

## Sources

1.  Creating a Table (CREATE TABLE) Chapter & Page Range: Chapter 1 (Creating Your First Database and Table), pages 5–7.

2.  Inserting Data (INSERT INTO) Chapter & Page Range: Chapter 1 (Creating Your First Database and Table), pages 8–9.

3.  Richard On Data. (2021, January 28). *Designing tables in R with "knitr" and "kableExtra" \| R Tutorial (2021)* \[Video\]. YouTube. <https://www.youtube.com/watch?v=JqUViTDoSEo>
