---
title: "Chess Tournament"
author: "David Melchor"
format: 
  html:
    embed-resources: true
    toc: true
---

## Introduction

In this project, I've been given a text file with chess tournament results where the information has some structure. My job is to create an R Markdown file that generates a .csv file that could be imported into a SQL database with the following information for all of the players:

- Player Name
- Player State
- Total Number of Points
- Player Pre-Rating
- Average Pre Chess Rating of Opponents

The **Average Pre Chess Rating of Opponents** was calculated by using the pre-tournament opponents’ ratings of 1436, 1563, 1600, 1610, 1649, 1663, 1716, and dividing by the total number of games played.

## Loading Packages
```{r packages}

# Load packages
pacman::p_load(tidyverse, stringr)
```

## Import Data
```{r data-import}

# URL
url <- "https://raw.githubusercontent.com/Dave-Melchor/Data-607-Data-Acquisition-and-Management/main/data/raw/tournamentinfo.txt"

# Use URL to bring in .txt data
raw_data <- readLines(url)

# Inspect the data
head(raw_data, 10)
```

Looking at the data, there are a lot of lines and space that doesn't include the information that we are trying to get. Each player data is found in two lines. For example, player 1's data is on line 5 and line 6 and player 2's data in found on line 8 and line 9 so we have to trim all that extra space.

## Data Cleaning
```{r cleaning}

# Generate row indices for Line 1 (Names, Points, Opponents)
line1_indices <- seq(5, length(raw_data), by = 3)

# Generate row indices for Line 2 (State, Ratings)
line2_indices <- seq(6, length(raw_data), by = 3)

# Extract those subset vectors
player_lines1 <- raw_data[line1_indices]
player_lines2 <- raw_data[line2_indices]

# Inspect the first 2 elements of each vector
head(player_lines1, 2)
head(player_lines2, 2)
```

Now we have two vectors where index 1 in both vectors corresponds to player 1 and index 2 in both vectors corresponds to player 2.

## Extracting Player Names
```{r get-names}

# Split each line by the vertical pipe
split_line1 <- str_split(player_lines1, "\\|")

# Extract the 2nd piece (Name) from every split line, and trim extra white space
player_names <- str_trim(sapply(split_line1, `[`, 2))

# Inspect the first 5 names
head(player_names, 5)
```

